Software Development in Startup Companies: The Green-field Startup Model

### 2013-02-13 article/main.tex
adjusted the complete methodology figure

The initial design of the case study is supported by the results of the \textit{systematic mapping study} (SMS) \cite{SMS} which contributed to define the initial questions for both questionnaires and semistructured interviews. The process depicted in Figure \ref{fig:gt:completemethodology} is evolutionary and affects the design at each new iteration.  In \textit{data collection} the empirical results are stored in a single \textit{case study database} and subsequently processed in \textit{data analysis} to form the theoretical categories.  At each iteration the new emergent theory is updated following a formal procedure (\textit{Theory generation}), and after verifying that the \textit{theoretical saturation}\footnote{The point at which executing more interviews wouldn't bring any additional value for constructing the theory.} of categories has been achieved, we finally proceeded to \textit{theory validation} or performed another complete cycle.

@@

Finally, startups present founders-centric structures, and especially in the early-stage, it is clear that \textit{CTO/CEO background has high-impact} on the company's development approach. For instance, in case of an academic background, the CTO encourages the introduction of some architectural design before the development phase. Moreover, even though the CTO/CEO starts guiding initially the development process, most of the decision are taken consensually by all members of the team. Then, the CTO/CEO decides only in particular situations where some conflicts occur.
%--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------%
\subsection{Evolutionary approach}
\label{res:gsm:cat2}
Among different approaches to development, startups prefer to build an initial prototype and iteratively refine it over time, similarly to the concept of \textit{``evolutionary prototyping''} \ref{EvProt}.  The reason of using this approach is that startups want to validate the product against the market as soon as possible, finding the proper product/market fit. Indeed, they can focus on developing only parts of the system they want to validate instead of working on developing a whole new system. Then, as the prototype is released, users detect opportunities for new functionalities and improvements, providing their feedback to developers.

@@

Instead of traditional requirement engineering activities, startups make use of \textit{informal specification of functionalities} by means of ticked-based tools to manage low-precision lists of features to implement, written in form of self-explanatory user stories \ref{AgilePlan}. Practitioners intensively use physical tools such as post-it notes and whiteboards, which help in making functionalities visible and prioritizing stories based on personal experiences. C4 comments \textit{``[\ldots] it is the only way. Too many people make the mistake of sitting down and write big specs and then they build it for four months, realizing the product is not valuable only at the end.''}

@@

Another important factor which contributes to \textit{accumulate the technical debt} is the general \textit{lack of architectural design}, substituted by \textit{high-level mockups and low-precision diagrams}, \textit{using modularization and frameworks from day 1} and \textit{describing critical interactions with third-party components only}. In particular, the use of well-known standards, frameworks and conventions limits the need of formal UML\ref{UML} diagrams and documentation, and provide a minimum level of maintainability costs. Additionally, since quality aspects are not a main concern (see \textit{limited number of functionalities} and {efficiency emerges after using the product} in CAT3) having a well-structured architecture remains a secondary priority, and design is conducted only when strictly necessary.

@@

In this subsection we present a list of the most relevant implications which emerge from the behaviour of early-stage startups, formally expressed in the model above, answering to RQ-2 (\textit{What is the current state-of-practice related to software development strategy in early-stage startups?}) through its sub-questions. Although the startups we inquired in the case study were spread across different nations and market sectors, we have found several patterns that emerged very clearly:

@@

\item Startups do not apply any standard development methodology. When focused on building the first version of the product, startups do not observe any specific and standard development methodologies or processes\footnote{Is worth noticing however that when asked the first question \textit{``Did you use any specific development methodology?''} 10 startups out of 13 referred to some kind of customized light version of SCRUM or XP (see Subsection \ref{sect:res:gt:meth-compare} for more details).}.  From an accurate and in-depth investigation into day-by-day activities, it emerged that the number of  the observed development \textit{practices} is extremely small and never systematic (only little \textit{pair programming} sessions and rarely \textit{test driven development}). Despite the wide number of publications, which discuss the adoption of lightweight methodologies, we have found that especially in this first stage not only methodologies but even individual practices are basically neglected.

@@

\item The greatest part of engineering activities of startups are focused on the implementation while only little attention is given to more conventional activities (project management, requirement specifications, analysis, architecture design, automatic testing, \ldots). Especially when bringing the first product to the market, the attitude of startups towards traditional engineering activities is strongly characterized by a general feeling of unnecessity. Every document, artifact or process, which does not produce a short-term tangible benefit to the software production is perceived as a waste of time. The most visible downside of this kind of approach is a growing \textit{accumulated technical debt} (Subsection \ref{sect:theory:debt}), which is partially mitigated by different elements supporting the development. Answering RQ-2.1, project management is substituted by the use of informal short-term deadlines supported by online collaborative tools with a kanban-board-like fashion; feature list are managed through ticket based systems supporting low-precision lists of requirement specifications in the form of user stories; after the essential set of functionality have been defined, making estimations and prioritization of features is not of much help; the traditional analysis process is, to a great extent, neglected (analysis is informally conducted for specific critical functionalities, usually by studying similar software solutions and rarely documenting decision-making processes); the design of the architecture is usually replaced by the use of well-known frameworks and naive diagrams on the whiteboard; advanced and modern version control systems work well for the configuration management; high-volumes of informal communication and the use of coding standards reduce the need of formal documentation; in the majority of startups writing automated test cases is perceived as a waste too (since the product reliability is not a primary concern,  engineers identify major bugs by testing the product internally and letting early adopters identify secondary bugs); finally, the infrastructural complexity of deployment is typically left to third party cloud applications.  Thus, by radically ignoring activities related to documentation, structures and processes, startups achieve a very high initial development speed in the first stage, but on the other hand they accumulate a technical debt that needs to be at least partially fulfilled when the company and the product grow.

\item The first release of the product includes only a limited set of well suitable functionalities focused on user experience. The beta version is usually focused on a minimal and essential set of functionalities providing a reasonable user experience (UX). As confirmed by practitioners, the user must be able to accomplish a task on the product without major interruptions in the flow, which would cause frustration before being able to capture his attention. To realize UX in a short time frame, developers focus on a limited number of suitable functionalities, especially in the first version(s) of the product. Other product qualities are regarded as less important\footnote{Note that from a business perspective, one might argue that established companies might take advantage of the new idea and develop a better version of the product internally. Despite this might represent a risk, established companies usually cannot afford to operate in a market of \textit{innovators} and \textit{early adopters} group of users - described by the Roger's bell curve in \cite{Rogers} - because of the little return of investment they would gain, as discussed by Christensen in \cite{Christensen2003}. Instead, established companies have more interests in acquiring a startup and all the knowledge it has gained during the \textit{customer development} process (see Appendix \ref{a:related:grey} for more details) while focusing effort on acquiring larger shares of a ``mass market'' sector.}: for instance, obtaining a high-reliability is not a priority since early adopters of innovative product are keen to accept minor bugs in exchange of new functionalities\footnote{These results further highlight differences between software development in startups and traditional web engineering, where quality aspects are rated as very important, even more than time-to-market \cite{Offutt2002}.}. Furthermore, thanks to cloud computing and more in general high availability of third party components, aspects such as performance, security and scalability can be achieved with a reasonable effort. Thus, given the general lack of quality concerns\footnote{The extent to which quality aspects are considered by the company, is strongly influenced by the market sector and the application type, discussed as confounding factors. However our case study has been carried out in startups working on innovative web application with very few quality concerns.}, and supported by the fact that the optimal strategy to improve the UX is by having the product used by real customers, startups can effectively focus on achieving short time-to-market.

@@

Additional validation procedures are presented in \cite{SMS}\footnote{The author performs a validation of the theoretical framework by triangulating the results of different methodologies with the existing literature. He highlights the areas which have been neglected by existing studies, providing possible directions for future studies. Finally he identifies and discusses confounding factors which could interfere with the theoretical model.}.
%--------------------------%%--------------------------%%--------------------------%%--------------------------%%--------------------------%%--------------------------%%--------------------------%
\subsection{Comparison with other frameworks}
\label{sect:theory:validation:others}

@@

Our theoretical framework explores the same challenges in software startups, where the act of tailoring processes leads developers to adopt only minimal practices, which are most suitable for the startups context. Furthermore, CTOs and CEOs' background has a great impact on the speed of the development process as described in Subsection \ref{sect:theory:team}. This is in accordance to Coleman's framework, which presents the background of founders and development managers as main factors that affect the management style of the development process.

@@

The author reports practitioners' perception that \textit{documentation} alone would not have ensured a complete shared understanding of project requirements. Moreover, defined processes are perceived by managers as additional items with a negative impact on the \textit{Creativity} and \textit{Flexibility} of the development team. This is in accordance with the generated theory (see Subsection \ref{sect:theory:theory}), which bases the reasons of adopting evolutionary and low-precision engineering elements to the flexibility and reactivity attributes of the development process.

@@

Also in \cite{Coleman2008}, the author describes how the \textit{``management approach''} is oriented towards \textit{``embrace and empower''} solutions in contrast with \textit{``command and control''} style, where there is an evidence of trust in development staff to carry out tasks with less direct supervision, greater delegation of responsibilities and a more generally consensual environment. Nevertheless, software development manager and founders have still impact on management style and indirectly on software development process. In the case of early-stage startups, founders are mainly software development managers as CEOs/CTOs and  technical practitioners at the same time. As Coleman identified the influence of the founders' and managers' background over the software development process, also GSM identifies how the  CEOs and CTOs background shapes the high-level strategies adopted in developing the initial product. Notwithstanding, team members remain self-organized, able to intervene in all the aspects of the development process without any direct supervision, as discussed in Subsection \ref{sect:theory:team}.

@@

An optimal combination of these three categories enables startups to ship code extremely quickly. However, the price that startups have to pay for achieving such high speed is related to the extent of the \textit{accumulated technical debt}, which will finally contribute to hinder performance in later stages. This is reflected by the right-hand side of the high-level framework, here extracted in Figure \ref{fig:an:comp:high-level2}.

>>>>

The initial design of the case study is supported by the results of the \textit{systematic mapping study} (SMS) \cite{SMS} which contributed to define the initial questions for both questionnaires and semistructured interviews. The process depicted in Figure \ref{fig:gt:completemethodology} is evolutionary and affects the design at each new iteration.  In \textit{data collection} the empirical results are stored in a single \textit{case study database} and subsequently processed in \textit{data analysis} to form the theoretical categories.  At each iteration the new emergent theory is updated following a formal procedure (\textit{Theory generation}), and after verifying that the \textit{theoretical saturation}\footnote{The point at which executing more interviews wouldn't bring any additional value for constructing the theory.} of categories has been achieved, we finally proceeded \textit{theory validation} or performed another complete cycle.

@@

Finally, startups present founders-centric structures, and especially in the early-stage, it is clear that \textit{CTO/CEO background has high-impact} on the company's development approach. For instance, in case of an academic background, the CTO encourages the introduction of some architectural design before the development phase. Moreover, even though the CTO/CEO starts guiding initially the development process, most of the decision are taken consensually by all members of the team. Then, the CTO/CEO decides only in particular situations where some conflicts occur.
%--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------%
\subsection{Evolutionary approach}
\label{res:gsm:cat2}
Among different approaches to development, startups prefer to build an initial prototype and iteratively refine it over time, similarly to the concept of \textit{``evolutionary prototyping''} \cite{EvProt}.  The reason of using this approach is that startups want to validate the product against the market as soon as possible, finding the proper product/market fit. Indeed, they can focus on developing only parts of the system they want to validate instead of working on developing a whole new system. Then, as the prototype is released, users detect opportunities for new functionalities and improvements, providing their feedback to developers.

@@

Instead of traditional requirement engineering activities, startups make use of \textit{informal specification of functionalities} by means of ticked-based tools to manage low-precision lists of features to implement, written in form of self-explanatory user stories \cite{AgilePlan}. Practitioners intensively use physical tools such as post-it notes and whiteboards, which help in making functionalities visible and prioritizing stories based on personal experiences. C4 comments \textit{``[\ldots] it is the only way. Too many people make the mistake of sitting down and write big specs and then they build it for four months, realizing the product is not valuable only at the end.''}

@@

Another important factor which contributes to \textit{accumulate the technical debt} is the general \textit{lack of architectural design}, substituted by \textit{high-level mockups and low-precision diagrams}, \textit{using modularization and frameworks from day 1} and \textit{describing critical interactions with third-party components only}. In particular, the use of well-known standards, frameworks and conventions limits the need of formal UML \cite{UML} diagrams and documentation, and provide a minimum level of maintainability costs. Additionally, since quality aspects are not a main concern (see \textit{limited number of functionalities} and {efficiency emerges after using the product} in CAT3) having a well-structured architecture remains a secondary priority, and design is conducted only when strictly necessary.

@@

In this subsection we present a list of the most relevant implications which emerge from the behaviour of early-stage startups, formally expressed in the GSM model, answering to our research question (\textit{What is the current state-of-practice related to software development strategy in early-stage startups?}) through its sub-questions. Although the startups we inquired in the case study were spread across different nations and market sectors, we have found several patterns that emerged very clearly:

@@

\item Startups do not apply any standard development methodology. When focused on building the first version of the product, startups do not observe any specific and standard development methodologies or processes.  From an accurate and in-depth investigation into daily activities, it emerged that the number of  the observed development \textit{practices} is extremely small and never systematic (only little \textit{pair programming} sessions and rarely \textit{test driven development}). Despite the wide number of publications, which discuss the adoption of lightweight methodologies, we have found that especially in this first stage not only methodologies but even individual practices are basically neglected.

@@

\item The greatest part of engineering activities of startups are focused on the implementation while only little attention is given to more conventional activities (project management, requirement specifications, analysis, architecture design, automatic testing, \ldots). Especially when bringing the first product to the market, the attitude of startups towards traditional engineering activities is strongly characterized by a general feeling of unnecessity. Every document, artifact or process, which does not produce a short-term tangible benefit to the software production is perceived as a waste of time. The most visible downside of this kind of approach is a growing \textit{accumulated technical debt} (CAT6), which is partially mitigated by different elements supporting the development. Project management is substituted by the use of informal short-term deadlines supported by online collaborative tools with a kanban-board-like fashion; feature list are managed through ticket based systems supporting low-precision lists of requirement specifications in the form of user stories; after the essential set of functionality have been defined, making estimations and prioritization of features is not of much help; the traditional analysis process is, to a great extent, neglected (analysis is informally conducted for specific critical functionalities, usually by studying similar software solutions and rarely documenting decision-making processes); the design of the architecture is usually replaced by the use of well-known frameworks and naive diagrams on the whiteboard; advanced and modern version control systems work well for the configuration management; high-volumes of informal communication and the use of coding standards reduce the need of formal documentation; in the majority of startups writing automated test cases is perceived as a waste too (since the product reliability is not a primary concern,  engineers identify major bugs by testing the product internally and letting early adopters identify secondary bugs); finally, the infrastructural complexity of deployment is typically left to third party cloud applications.  Thus, by radically ignoring activities related to documentation, structures and processes, startups achieve a very high initial development speed in the first stage, but on the other hand they accumulate a technical debt that needs to be at least partially fulfilled when the company and the product grow.

\item The first release of the product includes only a limited set of well suitable functionalities focused on user experience. The beta version is usually focused on a minimal and essential set of functionalities providing a reasonable user experience (UX). As confirmed by practitioners, the user must be able to accomplish a task on the product without major interruptions in the flow, which would cause frustration before being able to capture his attention. To realize UX in a short time frame, developers focus on a limited number of suitable functionalities, especially in the first version(s) of the product. Other product qualities are regarded as less important\footnote{Note that from a business perspective, one might argue that established companies might take advantage of the new idea and develop a better version of the product internally. Despite this might represent a risk, established companies usually cannot afford to operate in a market of \textit{innovators} and \textit{early adopters} group of users - described by the Roger's bell curve in \cite{Rogers} - because of the little return of investment they would gain, as discussed by Christensen in \cite{Christensen2003}. Instead, established companies have more interests in acquiring a startup and all the knowledge it has gained during the \textit{customer development} process while focusing effort on acquiring larger shares of a ``mass market'' sector.}: for instance, obtaining a high-reliability is not a priority since early adopters of innovative product are keen to accept minor bugs in exchange of new functionalities\footnote{These results further highlight differences between software development in startups and traditional web engineering, where quality aspects are rated as very important, even more than time-to-market \cite{Offutt2002}.}. Furthermore, thanks to cloud computing and more in general high availability of third party components, aspects such as performance, security and scalability can be achieved with a reasonable effort. Thus, given the general lack of quality concerns, and supported by the fact that the optimal strategy to improve the UX is by having the product used by real customers, startups can effectively focus on achieving short time-to-market.

@@

%--------------------------%%--------------------------%%--------------------------%%--------------------------%%--------------------------%%--------------------------%%--------------------------%
\subsection{Comparison with other frameworks}
\label{sect:theory:validation:others}

@@

Our theoretical framework explores the same challenges in software startups, where the act of tailoring processes leads developers to adopt only minimal practices, which are most suitable for the startups context. Furthermore, CTOs and CEOs' background has a great impact on the speed of the development process as described in CAT4. This is in accordance to Coleman's framework, which presents the background of founders and development managers as main factors that affect the management style of the development process.

@@

The author reports practitioners' perception that \textit{documentation} alone would not have ensured a complete shared understanding of project requirements. Moreover, defined processes are perceived by managers as additional items with a negative impact on the \textit{Creativity} and \textit{Flexibility} of the development team. This is in accordance with the generated theory, which bases the reasons of adopting evolutionary and low-precision engineering elements to the flexibility and reactivity attributes of the development process.

@@

Also in \cite{Coleman2008}, the author describes how the \textit{``management approach''} is oriented towards \textit{``embrace and empower''} solutions in contrast with \textit{``command and control''} style, where there is an evidence of trust in development staff to carry out tasks with less direct supervision, greater delegation of responsibilities and a more generally consensual environment. Nevertheless, software development manager and founders have still impact on management style and indirectly on software development process. In the case of early-stage startups, founders are mainly software development managers as CEOs/CTOs and  technical practitioners at the same time. As Coleman identified the influence of the founders' and managers' background over the software development process, also GSM identifies how the  CEOs and CTOs background shapes the high-level strategies adopted in developing the initial product. Notwithstanding, team members remain self-organized, able to intervene in all the aspects of the development process without any direct supervision, as discussed in CAT4.

@@

An optimal combination of these three categories enables startups to ship code extremely quickly. However, the price that startups have to pay for achieving such high speed is related to the extent of the \textit{accumulated technical debt}, which will finally contribute to hinder performance in later stages.
